---
title: "TBR01.1_EnvData_Investigation"
author: "Trine"
date: "2024-04-13"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE, 
                      warning=FALSE,
                      fig.align = "left",
                      fig.width = 10,
                      fig.height = 6)
```


```{r Libraries and seed}

# Libraries
library("tidyverse")       # data wrangling and visualisation   
library("corrplot")        # graphical display of a correlation matrix
library("GGally")          # ggplot2 extension for plots - inspect several plots together
library("vegan")           # multivariate analysis
library("patchwork")       # arrange plots
library ("cowplot")        # set a theme
library("here")            # set the path to the folder 
library("viridis")          # for color palettes
library("ggpubr")         # for arranging plots

set.seed(57)
```


```{r functions}

# A function that selects only the samples (rows) containing the specified environmental data. 
# This function makes it easier to deal with a dataset with a lot of lacking data for some samples.
envdata_available <- function(envdata, env_vector){

  
  # choose envfactors of interest
  envdata <- envdata[ , which(names(envdata) %in% env_vector)]
  # remove samples where env factors of interest are absent
  envdata <- envdata[complete.cases(envdata), ]
  #return cleaned datatabel to analyse
  return(envdata)
  

}

# A function that automatically generates a ttest with the samples that have data available for the specified environmental factors. \
# Returns a table with an overview of env factors tested, p-values, and the number of samples included in the ttest
ttest_function <- function(envdata, env_vector){
  # how to use:
  # envdata  = data.frame with environmental information
  # select_factors_ttest <- c("pH", "Temperature", "Conductivity", "Carbon", "Nitrogen", "CN")
  # p_val_df_Sample_type4 <- ttest_function(envdata = envdata, env_vector = select_factors_ttest)
  
    p_vals <- c()
    sig <- c()
    n_samples_included <- c()
    

  for (factor in select_factors_ttest){
    set.seed(57)
    select_factors <- c("Sample_type3", factor)
    test <- envdata_available(envdata=envdata, env_vector=select_factors)
    #test
    ttest_p <- t.test(test[, which(names(test) %in% c(factor))]~Sample_type3, data = test)$p.value
    p_vals <- c(p_vals, ttest_p)
    if (ttest_p < 0.05){
      sig <- c(sig, "significant")
    } else {
      sig <- c(sig, "not significant")
    }
    n_samples_included <- c(n_samples_included, nrow(test))
  }
  return(data.frame(select_factors_ttest, p_vals, sig, n_samples_included))
  

}


# Same function as before but removes outliers. 
# Returns a table with an overview of env factors tested, p-values, and the number of samples included in the ttest
ttest_function_no_outlier <- function(envdata, env_vector){
  # how to use:
  # envdata  = data.frame with environmental information
  # select_factors_ttest <- c("pH", "Temperature", "Conductivity", "Carbon", "Nitrogen", "CN")
  # p_val_df_Sample_type4 <- ttest_function(envdata = envdata, env_vector = select_factors_ttest)
  
  p_vals <- c()
  sig <- c()
  n_samples_included <- c()
  n_outliers <- c()
  
  for (factor in select_factors_ttest){
    set.seed(57)
    
    # which column no matches wanted - ie Temp = col 9
    select_factors <- c("Sample_type3", factor)
    # Get samples with the specified data available only
    test <- envdata_available(envdata=envdata, env_vector=select_factors)
    
  
    #Find outliers
    x <- test[, which(names(test) %in% c(factor))]
    Tmin = summary(x)[[2]]-(1.5*IQR(x))
    Tmax = summary(x)[[5]]+(1.5*IQR(x)) 
    outliers <- c(x[which(x < Tmin | x > Tmax)])
    
    test_sub <- data.frame(matrix(ncol=2,nrow=0))
    names(test_sub) <- names(test)
  
    #remove outliers before ttest
    for(i in 1:length(test[,1])){
      if (test[,2][i] < Tmin | test[,2][i] > Tmax){
        #print("Outlier")
        #print(test[,2][i])
      } else {
        j <- length(test_sub[,1]) + 1
        test_sub[j,] <- test[i,]
      }
    }
  
    #test
    ttest_p <- t.test(test_sub[, which(names(test_sub) %in% c(factor))]~Sample_type3, data = test_sub)$p.value
    p_vals <- c(p_vals, ttest_p)
    if (ttest_p < 0.05){
      sig <- c(sig, "significant")
    } else {
      sig <- c(sig, "not significant")
    }
    n_samples_included <- c(n_samples_included, nrow(test_sub))
    n_outliers <- c(n_outliers,  length(outliers)  )
    
  }
  
  return(data.frame(select_factors_ttest, p_vals, sig, n_samples_included, n_outliers))
}



# Make a themes for all the plots that look the same. 
# Use as normal themes: p + theme_boxplot()
theme_boxplot <- function(base_size=10, base_family="Courier"){
    #library(grid)
    library(ggthemes)
    (theme_bw(base_size=base_size, base_family=base_family)+
        theme(
        axis.title.x = element_blank (),
        axis.text.x  = element_blank(),
        axis.title.y = element_text(size=11),
        axis.text.y  = element_text(size=10),
        axis.ticks.x.bottom = element_blank(),
        strip.text = element_text(size=13),
        strip.background = element_blank(),
        #legend.position="none"
        ))
}

```


The goal of this analysis is to show that in terms of environmental data the sampling sites are different from each other - this enables the hypothesis that the microbial communities dwelling at these sites will be different as well

##### Overview of the data


##### Overview of the data
```{r load data, fig.height=13, fig.width=26}

#load data
envdata_all  <-  read.csv("phyloseq_files/DataEnv6.csv", header = TRUE, row.names = 3)

Na_upper_limit = 10000   
K_upper_limit = 10000
Al_upper_limit = 5000

envdata_all$Na <- as.numeric(replace(envdata_all$Na, envdata_all$Na == "HIGH", Na_upper_limit))
envdata_all$K <- as.numeric(replace(envdata_all$K, envdata_all$K == "HIGH", K_upper_limit))
envdata_all$Al <- as.numeric(replace(envdata_all$Al, envdata_all$Al == "HIGH", Al_upper_limit))
envdata_all

# Abiotic factors

# Create abiotic factors data table
envdata_all <- envdata_all %>% mutate(CN = Carbon / Nitrogen)
select_factors <- c("Name","Sample_type4","Sample_type4","Temperature","pH","Conductivity","Carbon","Nitrogen", "CN", "GWC")
abiotic <- envdata_available(envdata = envdata_all, env_vector = select_factors)

# Inspect the data for correlations between factors - using GGally


select_factors <- c("Sample_type4","Sample_type4","pH","Temperature","Conductivity","Carbon", "Nitrogen", "GWC",colnames(envdata_all[,15:39]))
icp_el <- envdata_available(envdata = envdata_all, env_vector = select_factors)

#Investigate how it affects the data that we have some upper detection limits for some of our samples with Na, K, and Al. 
icp_el %>%
  select(Sample_type4, Na, K, Al, Ca, Mg) %>%
  ggpairs(aes(color = Sample_type4))

#Correlating other minerals with Na, Al, and K, looks rather strange. Leave Na, K, and Al out of the analysis.

icp_el$Sample_type4

```


```{r load data, fig.height=13, fig.width=26}

#We cannot rely on the Na, K, and Al metals as these have reached a detection limit for some of the samples but not for other, presumably due to different running settings on the ICPMS machine.
#We therefore need to remove those from the dataset.
envdata_all <- envdata_all[ , -which(names(envdata_all) %in% c("Na", "K", "Al"))]


# Abiotic factors

# Create abiotic factors data table
envdata_all <- envdata_all %>% mutate(CN = Carbon / Nitrogen)
select_factors <- c("Name","Sample_type4","Sample_type4","Temperature","pH","Conductivity","Carbon","Nitrogen", "CN", "GWC")
abiotic <- envdata_available(envdata = envdata_all, env_vector = select_factors)



# 
# # Do any of the elements show correlations
select_factors <- c("Sample_type4","Sample_type4","pH","Temperature","Conductivity","Carbon", "Nitrogen", "GWC",colnames(envdata_all[,15:36]))
icp_el <- envdata_available(envdata = envdata_all, env_vector = select_factors)


icp_el2 <- icp_el
numeric_columns <- icp_el2[sapply(icp_el2, is.numeric)]

# Create an empty list to store results
correlation_results <- list()

# Loop through each pair of numerical columns
for (i in 1:(ncol(numeric_columns) - 1)) {
  for (j in (i + 1):ncol(numeric_columns)) {
    # Calculate correlation test
    corr_test <- cor.test(numeric_columns[, i], numeric_columns[, j], method = "spearman")
    
    # Store results in a list
    correlation_results <- c(correlation_results, list(
      c(colnames(numeric_columns)[i], colnames(numeric_columns)[j], corr_test$estimate, corr_test$p.value)
    ))
  }
}

# Convert the list to a data frame
correlation_df <- as.data.frame(do.call(rbind, correlation_results))
colnames(correlation_df) <- c("Variable_1", "Variable_2", "Correlation_Coefficient", "P_Value")

# Perform Benjamini-Hochberg adjustment
correlation_df$Adjusted_P_Value <- p.adjust(correlation_df$P_Value, method = "BH")
nrow(correlation_df)
#351 tests conducted

correlation_df2 <- correlation_df %>%
  dplyr::filter(Adjusted_P_Value < 0.05) %>%
  dplyr::filter(Correlation_Coefficient > 0.7) %>%
  dplyr::filter(Variable_1 != "Cd") %>%
  dplyr::filter(Variable_1 != "B") %>%
  dplyr::filter(Variable_1 != "Pb") %>%
  dplyr::filter(Variable_1 != "V") %>%
  dplyr::filter(Variable_1 != "Cr") %>%
  dplyr::filter(Variable_1 != "Mn") %>%
  dplyr::filter(Variable_1 != "Fe") %>%
  dplyr::filter(Variable_1 != "Co") %>%
  dplyr::filter(Variable_1 != "Cu") %>%
  dplyr::filter(Variable_1 != "Ba") %>%
  dplyr::filter(Variable_2 != "Cd")

correlation_df2$Correlation_Coefficient <- as.numeric(correlation_df2$Correlation_Coefficient)

###############
variable_names <- unique(c(correlation_df2$Variable_1, correlation_df2$Variable_2))

# Filter icp_el2 data frame to include only selected columns
icp_el3 <- icp_el2 %>%
  select(any_of(variable_names)) %>%
  cbind(., icp_el2$Sample_type4) %>%
  dplyr::rename("Sample_type4" = 9)

icp_tr <- icp_el3 %>%
  dplyr::filter(Sample_type4 == "Tramway Ridge")

numeric_columns <- icp_tr[sapply(icp_tr, is.numeric)]

correlation_results <- list()

# Loop through each pair of numerical columns
for (i in 1:(ncol(numeric_columns) - 1)) {
  for (j in (i + 1):ncol(numeric_columns)) {
    # Calculate correlation test
    corr_test <- cor.test(numeric_columns[, i], numeric_columns[, j], method = "spearman")
    
    # Store results in a list
    correlation_results <- c(correlation_results, list(
      c(colnames(numeric_columns)[i], colnames(numeric_columns)[j], corr_test$estimate, corr_test$p.value)
    ))
  }
}

# Convert the list to a data frame
correlation_df_tr <- as.data.frame(do.call(rbind, correlation_results))
colnames(correlation_df_tr) <- c("Variable_1", "Variable_2", "Correlation_Coefficient", "P_Value")

# Perform Benjamini-Hochberg adjustment
correlation_df_tr$Adjusted_P_Value <- p.adjust(correlation_df_tr$P_Value, method = "BH")

################
icp_ex <- icp_el3 %>%
  dplyr::filter(Sample_type4 == "Exp Hot Soil")

numeric_columns <- icp_ex[sapply(icp_ex, is.numeric)]

correlation_results <- list()

# Loop through each pair of numerical columns
for (i in 1:(ncol(numeric_columns) - 1)) {
  for (j in (i + 1):ncol(numeric_columns)) {
    # Calculate correlation test
    corr_test <- cor.test(numeric_columns[, i], numeric_columns[, j], method = "spearman")
    
    # Store results in a list
    correlation_results <- c(correlation_results, list(
      c(colnames(numeric_columns)[i], colnames(numeric_columns)[j], corr_test$estimate, corr_test$p.value)
    ))
  }
}

# Convert the list to a data frame
correlation_df_ex <- as.data.frame(do.call(rbind, correlation_results))
colnames(correlation_df_ex) <- c("Variable_1", "Variable_2", "Correlation_Coefficient", "P_Value")

# Perform Benjamini-Hochberg adjustment
correlation_df_ex$Adjusted_P_Value <- p.adjust(correlation_df_ex$P_Value, method = "BH")

##################
icp_sub <- icp_el3 %>%
  dplyr::filter(Sample_type4 == "Subglacial")

numeric_columns <- icp_sub[sapply(icp_sub, is.numeric)]

correlation_results <- list()

# Loop through each pair of numerical columns
for (i in 1:(ncol(numeric_columns) - 1)) {
  for (j in (i + 1):ncol(numeric_columns)) {
    # Calculate correlation test
    corr_test <- cor.test(numeric_columns[, i], numeric_columns[, j], method = "spearman")
    
    # Store results in a list
    correlation_results <- c(correlation_results, list(
      c(colnames(numeric_columns)[i], colnames(numeric_columns)[j], corr_test$estimate, corr_test$p.value)
    ))
  }
}

# Convert the list to a data frame
correlation_df_sub <- as.data.frame(do.call(rbind, correlation_results))
colnames(correlation_df_sub) <- c("Variable_1", "Variable_2", "Correlation_Coefficient", "P_Value")

# Perform Benjamini-Hochberg adjustment
correlation_df_sub$Adjusted_P_Value <- p.adjust(correlation_df_sub$P_Value, method = "BH")

##############
corr_all2 <- rbind(correlation_df2, correlation_df_tr, correlation_df_ex, correlation_df_sub)

####None of the correlations are significant within subcategory.

#write.csv(corr_all2, "correlations_sig_big.csv")
```




```{r, message=FALSE, fig.height=6.5, fig.width=8}
fig_cor_all <- icp_el %>%
  select(Sample_type4, Temperature, pH, Conductivity, Carbon, Nitrogen, GWC, Mg, Ca, Sr) %>%
  ggpairs(aes(color = Sample_type4))

fig_cor_env <- icp_el %>%
  select(Sample_type4, Temperature, pH, Conductivity, Carbon, Nitrogen) %>%
  ggpairs(aes(color = Sample_type4))

fig_cor_ions <- icp_el %>%
  select(Sample_type4, Mg, Ca, Sr) %>%
  ggpairs(aes(color = Sample_type4))


# ggsave("Saved_plots/fig_cor_env.png", 
#        fig_cor_env, 
#        width = 30, 
#        height = 20, 
#        units = "cm", 
#        dpi = 1000, 
#        device = "png")
# 
# ggsave("Saved_plots/fig_cor_ions.png", 
#        fig_cor_ions, 
#        width = 30, 
#        height = 20, 
#        units = "cm", 
#        dpi = 1000, 
#        device = "png")

```

There are a few interesting observations. For example that Ca and Sr seems to correlate and there are more of these in the subglacial samples compared to exposed.


##### Analysis by sampling site
```{r}

### stats for env parameters
#envdata_for_ttest <- envdata_all %>% mutate(CN = Carbon / Nitrogen)
#select_factors_ttest <- c("pH", "Temperature", "Conductivity", "Carbon", "Nitrogen", "CN", "GWC")

# Test if the levels of any of the chosen factors differ significantly between exposed and subglacial sites.
# Since some of the samples varies a lot and might have been sampled or tested differently, we want to check if outliers might skew the ttest.

#Without outliers removed:
#ttest_function(envdata = envdata_for_ttest, env_vector = select_factors_ttest)

#With outliers removed:
#ttest_function_no_outlier(envdata = envdata_for_ttest, env_vector = select_factors_ttest)
#When we remove outliers, CN is not significant anymore
```

```{r}

# Create a copy of the input dataframe with the same column and row names
env_data <- envdata_all[,9:38][,-4][,-28]
outlier_removed_data <- env_data

# Define the groups
groups <- c("Subglacial", "Tramway Ridge", "Exp Hot Soil")

# Loop through each group
for (group in groups) {
  # Subset the dataframe for the current group
  group_data <- env_data[env_data$group == group, ]
  
  # Loop through each column of the current group
  for (col in 1:ncol(group_data)) {
    # Exclude missing values from the column for outlier detection
    column_values <- group_data[, col]
    column_values_no_na <- column_values[!is.na(column_values)]
    
    # Calculate the lower and upper quartiles for the column
    lower_quartile <- quantile(column_values_no_na, 0.25, na.rm = TRUE)
    upper_quartile <- quantile(column_values_no_na, 0.75, na.rm = TRUE)
    
    # Calculate the interquartile range (IQR) for the column
    iqr <- upper_quartile - lower_quartile
    
    # Calculate the lower and upper bounds for outlier detection
    lower_bound <- lower_quartile - (1.5 * iqr)
    upper_bound <- upper_quartile + (1.5 * iqr)
    
    # Identify outliers and replace them with NA in the new dataframe
    outlier_removed_data[env_data$group == group, col] <- ifelse(column_values < lower_bound | column_values > upper_bound,
                                                                 NA, column_values)
  }
}

# Print the resulting table with outliers replaced by NAs and same column and row names
print(outlier_removed_data)

```




```{r, warning=FALSE}

### Analyse data by temperature in each site

outlier_removed_data$Sample_type4 <- envdata_all$Sample_type4

abiotic <- outlier_removed_data

#abiotic["Sample_type4"][abiotic["Sample_type4"] == "Exposed"] <- "Exposed"
#abiotic["Sample_type4"][abiotic["Sample_type4"] == "Subglacial"] <- "Subglacial"


# exposed_col <- "#F9A620"
# subglacial_col <- "#67a9cf"
# tr_col <- "#44AF69"

exposed_col <- "darkorange2"
subglacial_col <- "#386cb0"
tr_col <- "#7fc97f"



library("ggpubr")
library("EnvStats")
```




```{r}
### Metals
# # Start by selecting the metals we saw that may be different between transects
envdata_for_ttest <- envdata_all
#envdata_for_ttest <- outlier_removed_data

# outlier_removed_data
# envdata_all

select_factors_ttest <- c(colnames(envdata_all[16:36]))

p_vals_metals <- ttest_function_no_outlier(envdata = envdata_for_ttest, env_vector = select_factors_ttest)



# p values when outliers are taken out
p_vals_metals[p_vals_metals$sig == 'significant', ]

#p values when outliers are kept.
ttest_function(envdata = envdata_for_ttest, env_vector = select_factors_ttest)[p_vals_metals$sig == 'significant', ]


#choose the metals that significantly differ between exposed and subglacial sites
significant_Sample_type4 <- c(p_vals_metals[p_vals_metals$sig == 'significant', ]$select_factors_ttest)

# "Mg" "Ca" "Sr" "Ba" "S"  "Se" "Mn" "Fe" "Cd"
``` 

```{r}
library(FSA)
library(ggstatsplot)
abiotf <- abiotic %>%
  dplyr::select(29, 1:6, "Mg", "Ca","Sr", "S", "Se", "Cd")

abiotf[abiotf < 0.00000000] <- NA

#Replace
abiotf$Sample_type4 <- gsub("Subglacial", "3-Subglacial", abiotf$Sample_type4)

# Replace "Exp Hot Soil" with "2-Exp Hot Soil"
abiotf$Sample_type4 <- gsub("Exp Hot Soil", "2-Exp Hot Soil", abiotf$Sample_type4)

# Replace "Tramway Ridge" with "1-Tramway Ridge"
abiotf$Sample_type4 <- gsub("Tramway Ridge", "1-Tramway Ridge", abiotf$Sample_type4)

colnames_f <- colnames(abiotf)
plot_list <- list()

for(i in 2:length(colnames_f)) {
  plot_list[[i]] <- ggbetweenstats(data = abiotf, 
                                 x = "Sample_type4",
                                 y = !!colnames_f[i],
                                 plot.type = "box",
                                 type = "nonparametric",
                                 violin.args = list(color = "transparent",
                                                    fill = "transparent"),
                                 pairwise.comparisons = TRUE,
                                p.adjust.method  = "BH",
                                 xlab = "",
                                 results.subtitle = FALSE,
                                centrality.plotting = FALSE,
                                mean.point.args = list(color = "transparent",
                                                       fill = "transparent"),
                      point.args = list(alpha = 1, size = 3,
                                        position = ggplot2::position_jitterdodge(dodge.width = 0.6),
                                        stroke = 0, na.rm = TRUE),
                      ggtheme = theme_bw()) +
    scale_color_manual(values = c(tr_col, exposed_col, subglacial_col)) +
    theme(axis.text.x  = element_blank(),
          axis.ticks.x = element_blank(),
    axis.text.y  = element_text(size=10),
    axis.title.y = element_text(size=12),
    axis.title.y.right = element_blank(), 
    axis.text.y.right = element_blank(), 
    axis.ticks.y.right = element_blank()) + 
      stat_n_text(size=3, ypos = 3)
}

plot_list <- plot_list[-1]

env_plotsf <- ggarrange(plotlist = plot_list, 
                        common.legend = TRUE,
                        ncol = 3, nrow = 4)
env_plotsf

#ggsave("Figure 2_2.svg", env_plotsf, width = 240, height = 300, units = "mm", dpi = 300)


test <- ggbetweenstats(data = abiotf, 
                                 x = "Sample_type4",
                                 y = "Temperature",
                                 plot.type = "box",
                                 type = "nonparametric",
                                 violin.args = list(color = "transparent",
                                                    fill = "transparent"),
                                 pairwise.comparisons = TRUE,
                                p.adjust.method  = "BH",
                                 xlab = "",
                                 results.subtitle = FALSE,
                                centrality.plotting = FALSE,
                                mean.point.args = list(color = "transparent",
                                                       fill = "transparent"),
                      point.args = list(alpha = 1, size = 3,
                                        position = ggplot2::position_jitterdodge(dodge.width = 0.6),
                                        stroke = 0, na.rm = TRUE),
                      ggtheme = theme_bw()) +
    scale_color_manual(values = c(tr_col, exposed_col, subglacial_col)) +
    theme(axis.text.x  = element_blank(),
          axis.ticks.x = element_blank(),
    axis.text.y  = element_text(size=10),
    axis.title.y = element_text(size=12),
    axis.title.y.right = element_blank(), 
    axis.text.y.right = element_blank(), 
    axis.ticks.y.right = element_blank()) + 
      stat_n_text(size=3, ypos = 3)

test

####################

medians <- abiotf %>%
  dplyr::group_by(Sample_type4 ) %>%
  dplyr::summarise(across(everything(), median, na.rm = TRUE))

medians
```