---
title: "01_Alpha_Beta"
author: "Trine"
date: "2023-01-11"
output: html_document
---

```{r libraries}
library("ggplot2")
library("phyloseq")
library(here)
library(vegan) #to test which groupings that makes sense to make
```

```{r read_phyloseq}

# WHEN YOU OPEN THIS FILE ----
phy_all <- readRDS("phyloseq_files/physeq_all")

#otu_table(phy_all) # it still has the seqs instead of ASV_1, ASV_2, etc...

new.names <- paste0("ASV_", seq(ntaxa(phy_all))) # define new names ASV_1, ASV_2, ...
seqs <- taxa_names(phy_all) # store sequences
names(seqs) <- new.names # make map from ASV to full sequence
taxa_names(phy_all) <- new.names # rename 
phy_all


initial_ntaxa <- ntaxa(phy_all)

#Clean 
phy_all <- subset_taxa (phy_all, (Kingdom !="Eukaryota"))
phy_all <- subset_taxa (phy_all, (Order !="Chloroplast") | is.na(Order))    
phy_all <- subset_taxa (phy_all, (Family != "Mitochondria") | is.na(Family))
cat("\n")
cat("\nInitial number of taxa: ", initial_ntaxa)
cat("\nNumber of taxa excluding eukaryotes, chloroplast and mitochondrial seqs: ", ntaxa(phy_all))
cat("\nDifference: ", initial_ntaxa - ntaxa(phy_all))
phy_all

```

## Alpha Diversity


```{r AlphaDiv boxplot, warning=FALSE,fig.height=6, fig.width=12}

#exposed_col <- "#F9A620"
#subglacial_col <- "#67a9cf"
subcat_pal <- c("darkorange2","#386cb0","#7fc97f")


p_richness_box <-  
  plot_richness(phy_all, x="Sample_type4", measures=c("Shannon")) + 
  geom_boxplot(aes(fill=Sample_type4), alpha=0.8) +    
  labs(y="Shannon Index")+
  facet_grid(~Sample_type3, scales = "free", space = "free") +
  
  theme_bw()+
  
  theme(text= element_text(size=20),
        axis.text.x = element_text(size=20, angle =0 , hjust = 0.5, vjust = 0), 
        axis.title.x = element_text(size = 20),
        axis.title.x.bottom = element_blank(),
        axis.text.y = element_text(size = 20), axis.title.y = element_text(size=20),
        strip.text.x = element_text(size = 19),
        legend.text = element_text(size=25), 
        legend.title = element_text(size=30), 
        panel.background = element_blank(),
        strip.background = element_blank(),
        legend.position="none",
        ) +
  scale_fill_manual(values = subcat_pal)

p_richness_box$layers <- p_richness_box$layers[-1]

p_richness_box_save <- p_richness_box + 
  geom_point(size = 3, position = position_jitter(width = 0.2), color= "#333333") +
  theme(axis.text.x = element_blank())
                          
p_richness_box_save
```

## Beta Diversity
```{r,fig.height= 9, fig.width= 15, warning=FALSE}

phy_all_t <- microbiome::transform(phy_all, "clr")  #investigate what this does
ordu_subs = ordinate(phy_all_t, "PCoA", "unifrac")

p_beta <- plot_ordination(phy_all_t, ordu_subs, color="Sample_type4", shape = "Sample_type3") +
          geom_point(size=14, alpha=0.7) +
          guides(shape = guide_legend(order = 1))+ 
          labs(shape = "Category", colour = "Subcategory")+
          theme_bw() +
          theme(legend.title = element_text(size=17, hjust = 0), 
                legend.text = element_text(size = 15),
                axis.title.x = element_text(size = 20), 
                axis.title.y = element_text(size = 20),
                axis.text.x = element_text(size=13),
                axis.text.y = element_text(size=13),
                legend.key.size = unit(0.4, "cm"), 
                legend.position = c(-0.35,0.175))


p_beta_save <- p_beta + scale_color_manual(values=subcat_pal)


p_beta + scale_color_manual(values=subcat_pal)+ geom_text(mapping = aes(label = Name), size = 8, vjust=1.5)
p_beta_save

```


```{r,fig.height= 8, fig.width= 14}

ggplot_empt <- ggplot() + theme_classic() + theme(axis.line = element_line(colour = "white"))

pp_alpha1 <- ggarrange(p_richness_box_save, ggplot_empt, 
                       ncol =1, 
                       nrow=2, 
                       heights = c(1, 0.9))
pp_alpha1

fig_alpha_beta <- ggarrange(pp_alpha1, p_beta_save, 
          ncol =2, 
          nrow=1, 
          widths = c(0.5, 1))
fig_alpha_beta

ggsave("Saved_plots/TBR02_fig_alpha_beta.pdf",
       fig_alpha_beta,
       width = 14,
       height = 8,
       device = "pdf")

```


STEPHEN1: Delete this below? Or is it cool to have? The two chunks below are just exploring.

### Investigating which ordination we like
```{r exploring ordinations, warning=FALSE, message=FALSE, fig.height=13, fig.width=20}

library("plyr"); packageVersion("plyr")

#prune taxa and see if it influences the ordination plots.
topASVnames <- names(sort(taxa_sums(phy_all), decreasing = TRUE))[1:1000]  # -> if you want to look at only top 20, add [1:20]
phy_all_top <- prune_taxa(topASVnames, phy_all) #ASV.top200 is now a phyloseq with the 200 most


#phy_all_top <- phy_all
dist = "unifrac"
ord_meths = c("DCA", "CCA", "RDA", "DPCoA", "NMDS", "MDS", "PCoA")
plist = llply(as.list(ord_meths), function(i, phy_all_top, dist){
        ordi = ordinate(phy_all_top, method=i, distance=dist)
        plot_ordination(phy_all_top, ordi, "Name", color="Sample_type4",label="Name")
        }, phy_all_top, dist)

names(plist) <- ord_meths

pdataframe = ldply(plist, function(x){
  df = x$data[, 1:2]
  colnames(df) = c("Axis_1", "Axis_2")
  return(cbind(df, x$data))
  })

names(pdataframe)[1] = "method"
pdataframe

p = ggplot(pdataframe, aes(Axis_1, Axis_2, color=Sample_type4, label=Name)) + geom_point(size=5, alpha=0.5) + 
  facet_wrap(~method, scales="free")+ #scale_colour_gradient(low = "navy", high = "red")+ 
  theme_bw()+geom_text(aes(label=Name),hjust=0, vjust=0, size=5)+ 
  ggtitle("Distance: unifrac")
p

p_sub_exp = ggplot(pdataframe, aes(Axis_1, Axis_2, color=Sample_type3, label=Name)) + geom_point(size=5, alpha=0.5) + facet_wrap(~method, scales="free")+ #scale_colour_gradient(low = "navy", high = "red")+ 
  theme_bw()+geom_text(aes(label=Name),hjust=0, vjust=0, size=5)+ 
  ggtitle("Distance: unifrac")
p_sub_exp

p_ord_pH = ggplot(pdataframe, aes(Axis_1, Axis_2, color=pH, label=Name)) + geom_point(size=5, alpha=0.5) + facet_wrap(~method, scales="free")+ #scale_colour_gradient(low = "navy", high = "red")+ 
  scale_colour_gradient(low = "green4", high = "gold")+ theme(legend.title = element_text(size=10))+
  theme_bw()+geom_text(aes(label=Name),hjust=0, vjust=0, size=5)+ 
  ggtitle("Distance: unifrac")
p_ord_pH

p_ord_temp = ggplot(pdataframe, aes(Axis_1, Axis_2, color=Temperature, label=Name)) + geom_point(size=5, alpha=0.5) + facet_wrap(~method, scales="free")+ #scale_colour_gradient(low = "navy", high = "red")+ 
  scale_colour_gradient(low = "blue", high = "red")+ theme(legend.title = element_text(size=10))+
  theme_bw()+geom_text(aes(label=Name),hjust=0, vjust=0, size=5)+ 
  ggtitle("Distance: unifrac")
p_ord_temp

```

```{r,fig.height= 13, fig.width= 20}

theme_beta <- theme(legend.title = element_text(size=30, hjust = 0), 
                    legend.text = element_text(size = 25),
                    axis.title.x = element_text(size = 20), 
                    axis.title.y = element_text(size = 20),
                    axis.text.x = element_text(size=15),
                    axis.text.y = element_text(size=15),
                    legend.key.size = unit(0.99, "cm"))
subcat_shapes <- c(18,19,17,15,16)


p_beta_ph <- plot_ordination(phy_all_t, ordu_subs, color="pH", shape = "Sample_type4") +
                            geom_point(size=14, alpha=0.7) +
                            guides(shape = guide_legend(order = 1))+ 
                            labs(shape = "Subcategory", colour = "pH")+
                            theme_bw() + theme_beta + scale_colour_gradient(low = "green4", high = "gold") + scale_shape_manual(values=subcat_shapes)
p_beta_ph 
p_beta_ph + geom_text(mapping = aes(label = Name), size = 8, vjust=1.5)


p_beta_T <- plot_ordination(phy_all_t, ordu_subs, color="Temperature", shape = "Sample_type4") +
                            geom_point(size=14, alpha=0.7) +
                            guides(shape = guide_legend(order = 1))+ 
                            labs(shape = "Subcategory", colour = "Temperature (°C)")+
                            theme_bw() + theme_beta + scale_colour_gradient(low = "navy", high = "red") + scale_shape_manual(values=subcat_shapes)
p_beta_T + geom_text(mapping = aes(label = Name), size = 8, vjust=1.5)


p_beta_ca <- plot_ordination(phy_all_t, ordu_subs, color="Ca", shape = "Sample_type4") +
                              geom_point(size=14, alpha=0.7) +
                              guides(shape = guide_legend(order = 1))+ 
                              labs(shape = "Subcategory", colour = "Calcium (ppb)")+
                              theme_bw() + theme_beta + scale_colour_gradient(low = "purple4", high = "orange") + scale_shape_manual(values=subcat_shapes)
p_beta_ca + geom_text(mapping = aes(label = Name), size = 8, vjust=1.5)
```
## Add statistics


In this script we need to figure out how to divide our samples into groups.
Show that the groups you choose differ significantly in microbial composition. 

```{r}
#filter and transform data

erebusf <- phy_all
erebusfr <- microbiome::transform(erebusf, "clr")
unifrac.erebusfr <- phyloseq::distance(erebusfr, method = "unifrac")   #WHAT is the difference of distance and ordinate???

```

#### Environmental & Biological data analysis
```{r}

# From the beta diversity plot, it looks like several types of sites might be significantly different based on the microbial communities:
# Exposed and subglacial sites seems to differ, but also smaller groupings like Cave, Circums, Tramway, Crater, Westerb Crater.

# Now we will perform a PERMANOVA to see if these groups of sites have different centroids
# This will be complemented with a Permutation test for homogeneity of multivariate dispersions 
# -if this doesn't come back significant we cam trust that the PERMANOVA results show
# differences in data centroids and not in data dispersion

# Create a data frame with the environmental data 
envdata <- data.frame(sample_data(erebusfr))
erebusf <- phy_all
erebusfr <- microbiome::transform(erebusf, "clr")
#erebusfr<- transform_sample_counts(erebusf, function(x) x / sum(x) )
unifrac.erebusfr <- phyloseq::distance(erebusfr, method = "unifrac")
#unifrac.erebusfr <- phyloseq::distance(erebusfr, method = "wunifrac")


#test with unifrac (doesnt take the microbial community into account)

# PERMANOVA on Sample_type4: Exp hot soil, Tramway, Subglacial
unique(envdata$Sample_type4) # 
adonis2(unifrac.erebusfr ~ Sample_type4, data = envdata) # 0.001 ***
disp.Transect<- betadisper(unifrac.erebusfr, envdata$Sample_type4) 
permutest(disp.Transect) # 0.043 *

# PERMANOVA on "Subglacial" "Exposed" 
unique(envdata$Sample_type3) 
adonis2(unifrac.erebusfr ~ Sample_type3, data = envdata) # 0.001 ***
disp.site <- betadisper(unifrac.erebusfr, envdata$Sample_type3) # 0.024 *
permutest(disp.site)

# PERMANOVA on "Cave"           "Circum"         "Tramway"        "Crater"         "Western_Crater" "Mat"
unique(envdata$Sample_type2) # 
adonis2(unifrac.erebusfr ~ Sample_type2, data = envdata) # 0.001 ***
disp.Transect<- betadisper(unifrac.erebusfr, envdata$Sample_type2) 
permutest(disp.Transect) # 0.758


# PERMANOVA on "Other"          "Tramway"        "Western_Crater"
unique(envdata$Sample_type1) # 
adonis2(unifrac.erebusfr ~ Sample_type1, data = envdata) # 0.001 ***
disp.Transect<- betadisper(unifrac.erebusfr, envdata$Sample_type1) 
permutest(disp.Transect) #0.001 ***
# Tramway Ridge, Western Crater, and Other, have significantly differently dispersed values.

# PERMANOVA on "Other"   "Tramway" , Should I just make a ttest here?
unique(envdata$Sample_type0) # 
adonis2(unifrac.erebusfr ~ Sample_type0, data = envdata) # 0.001 ***
disp.Transect<- betadisper(unifrac.erebusfr, envdata$Sample_type0)
permutest(disp.Transect) # 0.001 *** 
#Tramway and Other have significantly differently dispersed values. 


#STEPHEN2: With pairwaise adonis, the different categories do look different, but since we concluded that they have dispersed values, I doesnt count or?
# Should I just delete the code below?

pairwise.adonis(unifrac.erebusfr, envdata$Sample_type3)
pairwise.adonis(unifrac.erebusfr, envdata$Sample_type4)
pairwise.adonis(unifrac.erebusfr, envdata$Sample_type0)



```
