---
title: "Class abundance"
author: "Trine"
date: "2023-07-05"
output: html_document
---

## Analysis of microbial community
```{r}

# Libraries
library("tidyverse")       # data wrangling and visualisation
library("phyloseq")        # analysis of microbial communities
library("patchwork")       # plots together
library("here")            # set the path to the folder
library("ggplot2")
library("ggpubr") #for putting figs together
library("RColorBrewer") #for color palettes
library("svglite") #for saving SVGs
library(here)
library("svglite") #for saving SVGs

set.seed(57)
```


```{r read_phyloseq}
# WHEN YOU OPEN THIS FILE ----
phy_all <- readRDS("phyloseq_files/physeq_all")

#otu_table(phy_all) # it still has the seqs instead of ASV_1, ASV_2, etc...

new.names <- paste0("ASV_", seq(ntaxa(phy_all))) # define new names ASV_1, ASV_2, ...
seqs <- taxa_names(phy_all) # store sequences
names(seqs) <- new.names # make map from ASV to full sequence
taxa_names(phy_all) <- new.names # rename 
phy_all

initial_ntaxa <- ntaxa(phy_all)

#Clean 
phy_all <- subset_taxa (phy_all, (Kingdom !="Eukaryota"))
phy_all <- subset_taxa (phy_all, (Order !="Chloroplast") | is.na(Order))    
phy_all <- subset_taxa (phy_all, (Family != "Mitochondria") | is.na(Family))
cat("\n")
cat("\nInitial number of taxa: ", initial_ntaxa)
cat("\nNumber of taxa excluding eukaryotes, chloroplast and mitochondrial seqs: ", ntaxa(phy_all))
cat("\nDifference: ", initial_ntaxa - ntaxa(phy_all))
```


#### Fig Stacked Barplot
```{r, fig.width=20, fig.height=11}
phy_all_trans <- transform_sample_counts(phy_all, function(x) x / sum(x))

# Data for plotting
erebus_p <- phy_all_trans %>% psmelt() 
```

```{r, fig.width=20, fig.height=11}
## Class

 pal_barnf <- c(
   "black",   "#9E0142", "#D53E4F", "#F46D43", "#FDAE61", "#FEE08B", "#FFFFBF", "#E6F598", "#ABDDA4", "#66C2A5", "#3288BD", "#5E4FA2",
 "#8E0152", "#C51B7D", "#DE77AE", "#F1B6DA", "#FDE0EF", "#E6F5D0", "#B8E186", "#7FBC41", "#4D9221", "#276419", "#40004B",
 "#762A83", "#9970AB", "#C2A5CF", "#E7D4E8", "#F7F7F7", "#D9F0D3", "#A6DBA0", "#5AAE61", "#1B7837", "#00441B", "#543005", "#8C510A",
 "#BF812D", "#DFC27D", "#F6E8C3", "#C7EAE5", "#80CDC1", "#35978F", "#01665E", "#003C30", "#636363")

erebus_p["Class"][which(erebus_p$Abundance < 0.02), ] <- "Class < 2 %"
erebus_p$Class[is.na(erebus_p$Class)] = "Unassigned"

x <- c("Class < 2 %",
"Acidimicrobiia","Acidobacteriae","Actinobacteria","AD3","Alphaproteobacteria", "Anaerolineae","Armatimonadia",
"Bacilli", "Bacteroidia","Blastocatellia","Chloroflexia","Chthonomonadetes","Cyanobacteriia","Deinococci","Desulfotomaculia",
 "FFCH5909","Fimbriimonadia","Gammaproteobacteria", "Gemmatimonadetes","Gitt-GS-136","Holophagae","KD4-96","Kryptonia",          
 "Ktedonobacteria","Myxococcia","Nitrososphaeria","Nitrospiria","Oligoflexia","P2-11E","Parcubacteria","Phycisphaerae",      
 "Planctomycetes","Polyangia","Sericytochromatia","SHA-26","Sulfobacillia","Thermoleophilia","TK10",     
 "Vampirivibrionia","Verrucomicrobiae","Vicinamibacteria","YNPFFP2","Unassigned")

names(pal_barnf) = x

erebus_p$Class <- factor(as.character(erebus_p$Class), levels=x)
erebus_p <- erebus_p[order(erebus_p$Class),]


# Plot 
plot_Class <- ggplot(erebus_p, aes(x = Name, y = Abundance, fill = Class)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = pal_barnf) +theme_bw()+
  theme(
    strip.background = element_blank(),
    strip.text.x = element_text(size = 9),
    panel.background = element_blank(),
    axis.ticks.x = element_blank(),
    axis.text.x  = element_text(size=9, vjust = 0.5, hjust = 1, angle = 90),
    axis.title.x = element_blank(),
    axis.text.y  = element_text(size=9),
    axis.title.y = element_text(size=10),
    plot.title = element_text(size=20, hjust = -0.05),
    legend.text = element_text(size=9), 
    legend.title = element_blank(),
    legend.position = "bottom",
    legend.key.width = unit(0.3, "cm"),
    legend.key.height=unit(0.3,"cm")) +
  coord_cartesian(expand = FALSE)



ploy_class_save <- plot_Class + facet_grid(~Sample_type4, scales = "free", space = "free")

#plot_Class$data$Sample_type3 <- factor(plot_Class$data$Sample_type3, levels = c("Subglacial", "Exposed"))
ploy_class_save$data$Sample_type4 <- factor(ploy_class_save$data$Sample_type4, levels = c("Subglacial", "Exp Hot Soil", "Western Crater", "Tramway Ridge"))


ploy_class_save <- ploy_class_save + 
                    theme(panel.margin=unit(.07, "lines"), 
                         panel.border = element_rect(color = "black", fill = NA, size = 1), 
                         strip.background = element_rect(color = "black", size = 1))

plot_Class_save_all <- ploy_class_save + scale_fill_manual(values = pal_barnf)
plot_Class_save_all
```


Add division into groups
```{r}

merged_group <- phyloseq::merge_samples(phy_all,"Sample_type4") # 5 samples
sample_data(merged_group)$Sample_type4 <- as.character(sample_names(merged_group))

temp <- as.data.frame(as.matrix(sample_data(merged_group))) %>%
  mutate(Sample_typeX = case_when(
    startsWith(Sample_type4, "Subglacial") ~ "Subglacial",
    startsWith(Sample_type4, "Tramway Ridge") ~ "Exposed",
    startsWith(Sample_type4, "Exp Hot Soil") ~ "Exposed",
    startsWith(Sample_type4, "Western Crater") ~ "Exposed",
    ))
sample_data(merged_group)$Sample_type3 <- temp$Sample_typeX

    
# Data for plotting - process the same way as Stephen did:
merged_group_trans <- transform_sample_counts(merged_group, function(x) x/sum(x))
merged_group_trans_p <- merged_group_trans %>% psmelt()


# reorder Class before naming the Class < 1% and the Unassigned. Otherwise it doesnt work for some reason.
merged_group_trans_p$Class <- factor(as.character(merged_group_trans_p$Class), levels=x)
merged_group_trans_p <- merged_group_trans_p[order(merged_group_trans_p$Class),]


length(merged_group_trans_p$Class[which(merged_group_trans_p$Class == "Class < 2 %")])
merged_group_trans_p["Class"][which(merged_group_trans_p$Abundance < 0.01), ] <- "Class < 2 %"
merged_group_trans_p$Class[is.na(merged_group_trans_p$Class)] = "Unassigned"

ggplot(merged_group_trans_p, aes(x = Sample_type4, y = Abundance, fill = Class)) +
  geom_bar(stat = "identity", width = 0.98) +
  scale_fill_manual(values = pal_barnf)


```


```{r,fig.width=20, fig.height=11}

# Plot 
plot_Class <- ggplot(merged_group_trans_p, aes(x = Sample_type4, y = Abundance, fill = Class)) +
  geom_bar(stat = "identity", width = 0.98) +
  scale_fill_manual(values = pal_barnf) + theme_bw()+
  theme(
    strip.background = element_blank(),
    strip.text.x = element_text(size = 10),
    panel.background = element_blank(),
    axis.ticks.x = element_blank(),
    legend.title = element_blank(),
    legend.text = element_text(size=9),
    legend.position = "none",
    axis.text.x  = element_text(size=9, vjust = 0.5, hjust = 0.5, angle = 0),
    axis.title.x = element_blank(),
    axis.text.y  = element_text(size=9),
    axis.title.y = element_text(size=10),
    plot.title = element_text(size=20, hjust = -0.05)) +
  coord_cartesian(expand = FALSE)

plot_Class$data$Sample_type4 <- factor(plot_Class$data$Sample_type4, levels = c("Subglacial", "Exp Hot Soil", "Western Crater", "Tramway Ridge"))
plot_Class$data$Sample_type3 <- factor(plot_Class$data$Sample_type3, levels = c("Subglacial", "Exposed"))

ploy_phylym_save <- plot_Class + facet_grid(~Sample_type3, scales = "free", space = "free") + 
  theme(panel.margin=unit(.05, "lines"),
        panel.border = element_rect(color = "black", fill = NA, size = 1.2), 
        strip.background = element_rect(color = "black", size = 1),
        legend.text = element_text(size=25), 
                                      legend.title = element_text(size=30))
#ploy_phylym_save

plot_Class_save_groups <- ploy_phylym_save

```


```{r, fig.width=13, fig.height=10}

class_abundance_comb <- (plot_Class_save_groups + theme(legend.position="none"))/ (plot_Class_save_all + theme(legend.position="bottom", 
                                                                                                         legend.text = element_text(size=15), 
                                                                                                         )) 
class_abundance2 <- ggarrange(plot_Class_save_groups,
                              plot_Class_save_all,
                              nrow = 2,
                              labels = c("A", "B"),
                              heights = c(0.45, 1))

class_all2 <- ggarrange(class_abundance2,
                      bps_c,
                      nrow = 2,
                      labels = c(""),
                      heights = c(1, 0.6))

class_all2

ggsave("barplots_class_all3.png", class_all2, path = "Saved_plots/",
       width = 240, height = 250, units = "mm", dpi = 300)
ggsave("barplots_class_all3.svg", class_all2, path = "Saved_plots/",
       width = 240, height = 250, units = "mm", dpi = 300)

```

```{r}
ggsave(
  "TBR03.2_class_abundance2.pdf",
  plot = class_abundance2,
  path = "Saved_plots/",
  device = "pdf",
  width = 13,
  height = 10,
)

```



# Exploring the data further

Investigate: which phyla does the unkown classes belong to?
```{r, fig.height=20, fig.width=30}
sample_data(phy_all_trans)
phy_class_na <- subset_taxa(phy_all_trans, is.na(Class))
tax_table(phy_class_na)

erebus_p_class_na <- phy_class_na %>% psmelt() 

plot_bar(phy_class_na, x="Name", y="Abundance", fill="Phylum")  + theme_bw()+
  theme(
    strip.background = element_blank(),
    strip.text.x = element_text(size = 16),
    panel.background = element_blank(),
    axis.ticks.x = element_blank(),
    legend.title = element_blank(),
    legend.text = element_text(size=12),
    axis.text.x  = element_text(size=16, vjust = 0.5, hjust = 1, angle = 90),
    axis.title.x = element_blank(),
    axis.text.y  = element_text(size=16),
    axis.title.y = element_text(size=16),
    plot.title = element_text(size=20, hjust = -0.05)) +
  coord_cartesian(expand = FALSE)+
  facet_grid(~Sample_type4, scales = "free", space = "free") + 
  theme(panel.margin=unit(.05, "lines"),
        panel.border = element_rect(color = "black", fill = NA, size = 1.2), 
        strip.background = element_rect(color = "black", size = 1),
        legend.text = element_text(size=25), 
        legend.title = element_text(size=30))

```
Investigate: What are the abundant ASVs of unkown classes?
```{r,fig.height=20, fig.width=30}
erebus_p_class_na["OTU"][which(erebus_p_class_na$Abundance < 0.01), ] <- "Class < 1 %"

# Plot 
plot_class_na <- ggplot(erebus_p_class_na, aes(x = Name, y = Abundance, fill = OTU)) +
  geom_bar(stat = "identity") +theme_bw()+
  theme(
    strip.background = element_blank(),
    strip.text.x = element_text(size = 16),
    panel.background = element_blank(),
    axis.ticks.x = element_blank(),
    axis.text.x  = element_text(size=16, vjust = 0.5, hjust = 1, angle = 90),
    axis.title.x = element_blank(),
    axis.text.y  = element_text(size=16),
    axis.title.y = element_text(size=16),
    plot.title = element_text(size=16, hjust = -0.05),
    legend.text = element_text(size=16), 
    legend.title = element_text(size=30, angle = 90),
    legend.position = "right") +
  coord_cartesian(expand = FALSE) +
  scale_fill_manual(values = c(
   "black",   "#9E0142", "#D53E4F", "#F46D43", "#FDAE61", "#FEE08B", "#ABDDA4", "#66C2A5", "#3288BD", "#5E4FA2",
 "#8E0152", "#C51B7D", "#DE77AE", "#F1B6DA", "#7FBC41", "#4D9221", "#276419", "#40004B",
 "#762A83", "#9970AB", "#C2A5CF", "#E7D4E8", "#A6DBA0", "blue", "red", "#00441B", "#543005", "#8C510A",
 "#BF812D", "#DFC27D", "#80CDC1", "#35978F", "#01665E", "#003C30", "#636363"))

plot_class_na_save <- plot_class_na + facet_grid(~Sample_type4, scales = "free", space = "free")

#plot_Class$data$Sample_type3 <- factor(plot_Class$data$Sample_type3, levels = c("Subglacial", "Exposed"))
plot_class_na_save$data$Sample_type4 <- factor(plot_class_na_save$data$Sample_type4, levels = c("Subglacial", "Exp Hot Soil", "Western Crater", "Tramway Ridge"))

plot_class_na_save
```
I BLASTed some of these ASVs:
```{r}
seqs["ASV_17"]
#This ASV is found in other mountain environments, Antarctica, and acidic mining evironments!
# 100% identity but is this published? FQ716929
#HF930837, KJ623632, HF546501, JX505051 (!!), FR749815, FN870208, JF737898, DQ660863 , MK724119, MK242442, MK521747

seqs["ASV_16"]
#ONLY 95.8% identity!
# looks like the closest relatives are mainly found in soil samples from forests?? This is not expected. Maybe a Firmicute: GU016168.1 95% identity
# This article keeps coming up: Ureolytic prokaryotes in soil; community abundance and diversity

seqs["ASV_8"]
# Several results with 100% identity. But what is it?
#Glacial movement shapes bacterial community structure (KX094452.1)
# Greenland? https://academic.oup.com/femsec/article/92/9/fiw127/2197710?login=false
#  KJ623633,, AB464941.2

seqs["ASV_142"]
#	94.42"

seqs["ASV_152"]
#	97.21%

seqs["ASV_118"]
#	97.21%

seqs["ASV_224"]
#	97.21%

```


Try to subset some interesting phyla that seem to differ between the sample categories
```{r}

phy_actino = subset_taxa(phy_all_trans, Phylum == "Actinobacteriota")

phy_chloro = subset_taxa(phy_all_trans, Phylum == "Chloroflexi")

phy_cyano = subset_taxa(phy_all_trans, Phylum == "Cyanobacteria")

phy_kingdom = subset_taxa(phy_all_trans, Kingdom == "Archaea")

```

```{r, fig.height=10, fig.width=14}

plot_bar(phy_actino, x="Name", y="Abundance", fill="Genus")  + theme_bw()+
  theme(
    strip.background = element_blank(),
    strip.text.x = element_text(size = 1),
    panel.background = element_blank(),
    axis.ticks.x = element_blank(),
    legend.title = element_blank(),
    legend.text = element_text(size=1),
    axis.text.x  = element_text(size=7, vjust = 0.5, hjust = 1, angle = 90),
    axis.title.x = element_blank(),
    axis.text.y  = element_text(size=1),
    axis.title.y = element_text(size=1),
    plot.title = element_text(size=2, hjust = -0.05)) +
  coord_cartesian(expand = FALSE)+
  facet_grid(~Sample_type4, scales = "free", space = "free") + 
  theme(panel.margin=unit(.05, "lines"),
        panel.border = element_rect(color = "black", fill = NA, size = 1.2), 
        strip.background = element_rect(color = "black", size = 1),
        legend.text = element_text(size=10), 
        legend.title = element_text(size=1)) #+ scale_fill_manual(values=c(Trine_palette, Trine_palette, "blue", "red", "black"))

erebus_p_phy_actino <- phy_actino %>% psmelt() 
erebus_p_phy_actino["OTU"][which(erebus_p_phy_actino$Abundance < 0.02), ] <- "Class < 2 %"

# Plot 
plot_actino <- ggplot(erebus_p_phy_actino, aes(x = Name, y = Abundance, fill = OTU)) +
  geom_bar(stat = "identity") +theme_bw()+
  theme(
    strip.background = element_blank(),
    strip.text.x = element_text(size = 16),
    panel.background = element_blank(),
    axis.ticks.x = element_blank(),
    axis.text.x  = element_text(size=16, vjust = 0.5, hjust = 1, angle = 90),
    axis.title.x = element_blank(),
    axis.text.y  = element_text(size=16),
    axis.title.y = element_text(size=16),
    plot.title = element_text(size=20, hjust = -0.05),
    legend.text = element_text(size=15), 
    legend.title = element_text(size=5, angle = 90),
    legend.position = "bottom") +
  coord_cartesian(expand = FALSE)

plot_actino <- plot_actino + facet_grid(~Sample_type4, scales = "free", space = "free")

#plot_Class$data$Sample_type3 <- factor(plot_Class$data$Sample_type3, levels = c("Subglacial", "Exposed"))
plot_actino$data$Sample_type4 <- factor(plot_actino$data$Sample_type4, levels = c("Subglacial", "Exp Hot Soil", "Western Crater", "Tramway Ridge"))

plot_actino

plot_bar(phy_chloro, x="Name", y="Abundance", fill="Class")  + theme_bw()+
  theme(
    strip.background = element_blank(),
    strip.text.x = element_text(size = 1),
    panel.background = element_blank(),
    axis.ticks.x = element_blank(),
    legend.title = element_blank(),
    legend.text = element_text(size=12),
    axis.text.x  = element_text(size=16, vjust = 0.5, hjust = 1, angle = 90),
    axis.title.x = element_blank(),
    axis.text.y  = element_text(size=16),
    axis.title.y = element_text(size=16),
    plot.title = element_text(size=20, hjust = -0.05)) +
  coord_cartesian(expand = FALSE)+
  facet_grid(~Sample_type4, scales = "free", space = "free") + 
  theme(panel.margin=unit(.05, "lines"),
        panel.border = element_rect(color = "black", fill = NA, size = 1.2), 
        strip.background = element_rect(color = "black", size = 1),
        legend.text = element_text(size=25), 
        legend.title = element_text(size=30))

plot_bar(phy_kingdom, x="Name", y="Abundance", fill="OTU")  + theme_bw()+
  theme(
    strip.background = element_blank(),
    strip.text.x = element_text(size = 1),
    panel.background = element_blank(),
    axis.ticks.x = element_blank(),
    legend.title = element_blank(),
    legend.text = element_text(size=12),
    axis.text.x  = element_text(size=16, vjust = 0.5, hjust = 1, angle = 90),
    axis.title.x = element_blank(),
    axis.text.y  = element_text(size=16),
    axis.title.y = element_text(size=16),
    plot.title = element_text(size=20, hjust = -0.05)) +
  coord_cartesian(expand = FALSE)+
  facet_grid(~Sample_type4, scales = "free", space = "free") + 
  theme(panel.margin=unit(.05, "lines"),
        panel.border = element_rect(color = "black", fill = NA, size = 1.2), 
        strip.background = element_rect(color = "black", size = 1),
        legend.text = element_text(size=25), 
        legend.title = element_text(size=30))



erebus_p_phy_king <- phy_kingdom %>% psmelt() 
erebus_p_phy_king["OTU"][which(erebus_p_phy_king$Abundance < 0.01), ] <- "Class < 1 %"

# Plot 
plot_king <- ggplot(erebus_p_phy_king, aes(x = Name, y = Abundance, fill = Order)) +
  geom_bar(stat = "identity") +theme_bw()+
  theme(
    strip.background = element_blank(),
    strip.text.x = element_text(size = 16),
    panel.background = element_blank(),
    axis.ticks.x = element_blank(),
    axis.text.x  = element_text(size=16, vjust = 0.5, hjust = 1, angle = 90),
    axis.title.x = element_blank(),
    axis.text.y  = element_text(size=16),
    axis.title.y = element_text(size=16),
    plot.title = element_text(size=20, hjust = -0.05),
    legend.text = element_text(size=15), 
    legend.title = element_text(size=5, angle = 90),
    legend.position = "bottom") +
  coord_cartesian(expand = FALSE)

plot_king <- plot_king + facet_grid(~Sample_type4, scales = "free", space = "free")

#plot_Class$data$Sample_type3 <- factor(plot_Class$data$Sample_type3, levels = c("Subglacial", "Exposed"))
plot_king$data$Sample_type4 <- factor(plot_king$data$Sample_type4, levels = c("Subglacial", "Exp Hot Soil", "Western Crater", "Tramway Ridge"))

plot_king

max(otu_table(phy_kingdom)[,"ASV_1"]) #TR1 - 32

plot_bar(phy_cyano, x="Name", y="Abundance", fill="Class")  + theme_bw()+
  theme(
    strip.background = element_blank(),
    strip.text.x = element_text(size = 16),
    panel.background = element_blank(),
    axis.ticks.x = element_blank(),
    legend.title = element_blank(),
    legend.text = element_text(size=12),
    axis.text.x  = element_text(size=16, vjust = 0.5, hjust = 1, angle = 90),
    axis.title.x = element_blank(),
    axis.text.y  = element_text(size=16),
    axis.title.y = element_text(size=16),
    plot.title = element_text(size=20, hjust = -0.05)) +
  coord_cartesian(expand = FALSE)+
  facet_grid(~Sample_type4, scales = "free", space = "free") + 
  theme(panel.margin=unit(.05, "lines"),
        panel.border = element_rect(color = "black", fill = NA, size = 1.2), 
        strip.background = element_rect(color = "black", size = 1),
        legend.text = element_text(size=25), 
        legend.title = element_text(size=30))

sample_data(phy_all)[,-1:-9]

```
Cyanobacteria in caves?: https://academic.oup.com/gbe/article/6/5/1031/603056
READ https://en.wikipedia.org/wiki/Vampirovibrio_chlorellavorus
https://www.science.org/doi/full/10.1126/science.aal3794


```{r}
otu_table(phy_cyano)[,"ASV_411"]
```





