---
title: "Microeco_heatmap"
author: "Trine"
date: "2023-03-26"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r Libraries and seed}

# Libraries
library("tidyverse")       # data wrangling and visualisation
library("phyloseq")        # analysis of microbial communities
library("vegan")           # ecological multivariate analyses
library("microbiome") #for analyzing phyloseq objects
library("viridis") #color package
library("svglite") #for saving SVGs
library("ggpubr") #for saving plots
library("here")            # set the path to the folder
library("ggplot2")
library("microeco")
library("file2meco")
library("agricolae")
library(cowplot)
library(here)

set.seed(57)
```




```{r read_phyloseq}

# WHEN YOU OPEN THIS FILE ----
phy_all <- readRDS("phyloseq_files/physeq_all")

#otu_table(phy_all) # it still has the seqs instead of ASV_1, ASV_2, etc...

new.names <- paste0("ASV_", seq(ntaxa(phy_all))) # define new names ASV_1, ASV_2, ...
seqs <- taxa_names(phy_all) # store sequences
names(seqs) <- new.names # make map from ASV to full sequence
taxa_names(phy_all) <- new.names # rename 
phy_all

initial_ntaxa <- ntaxa(phy_all)

#Clean 
phy_all <- subset_taxa (phy_all, (Kingdom !="Eukaryota"))
phy_all <- subset_taxa (phy_all, (Order !="Chloroplast") | is.na(Order))    
phy_all <- subset_taxa (phy_all, (Family != "Mitochondria") | is.na(Family))
cat("\n")
cat("\nInitial number of taxa: ", initial_ntaxa)
cat("\nNumber of taxa excluding eukaryotes, chloroplast and mitochondrial seqs: ", ntaxa(phy_all))
cat("\nDifference: ", initial_ntaxa - ntaxa(phy_all))

phy_all
```
```{r}
#Load version of physeq object, with contaminants removed but not normalized
volcsf <- phy_all
```


```{r}
#get all of the data ready to go into the microeco world
envdata <- data.frame(sample_data(volcsf))

#want to include ASV as a taxonomic rank
tax_table <- data.frame(tax_table(volcsf)) %>%
  rownames_to_column(., var = "ASV_original") %>%
  dplyr::mutate(., ASV = ASV_original) %>%
  column_to_rownames(., var = "ASV_original")
tax_table

#transform phyloseq to microeco object
volcsf_meco <- phyloseq2meco(volcsf)

volcsf_meco$sample_table <- envdata #if you need to modify the envdata as I did
volcsf_meco$tax_table <- tax_table
```


```{r, fig.height=8, fig.width=25}
pro.func <- trans_func$new(dataset = volcsf_meco)
pro.func$for_what <- "prok"
pro.func$cal_spe_func(prok_database = c("FAPROTAX"))

pro.func$cal_spe_func_perc(abundance_weighted = FALSE)

pro.func$res_spe_func_perc

save <- pro.func$plot_spe_func_perc() #+
  #facet_grid(~ pro.func$sample_table$Sample_type3, scales = "free")
```

```{r,fig.height=8, fig.width=25}
save 

#The percentages of the OTUs having the same trait can reflect the functional redundancy of this function in the community.

```



```{r}
# If you want to change the group list, reset the list t2$func_group_list
pro.func$func_group_list
pro.func$show_prok_func("methanotrophy")

```


```{r}
volcsf_meco$sample_table <- envdata #if you need to modify the envdata as I did
volcsf_meco$tax_table <- tax_table
```


Makes most sense when looking at sample type 3 (exposed/subglacial). Here it is obvious that the photosyntrophy differs between the groups which makes perfectly sense. Then also, the aerobic ammonia oxidation (probably the archaea in the exposed sites) and nitrification. Then also methaneotophy, methyltrophy, anaerobic chemoheterotrophy, hydrocarbon degradation, dark sulfide oxidation, nonphotosynthetic cyanobacteria!!!! Must be in the caves as I saw in my thesis. 


```{r, fig.height=6.5,fig.width=7}
envdata_f <- envdata[,c("Sample_type3", "Temperature", "pH", "Conductivity", "GWC", "Carbon", "Nitrogen", "Cd", "Mn", "Mg", "Ca", "Sr", "U", "Pb")]

envdata_f2 <- envdata_f %>%
  dplyr::mutate(Sample_type3 = ifelse(Sample_type3 == "Subglacial", 0, ifelse(Sample_type3 == "Exposed", 1, Sample_type3)))

envdata_f2$Sample_type3 <- as.numeric(envdata_f2$Sample_type3)

t3 <- trans_env$new(dataset = volcsf_meco, add_data = envdata_f2)
t3$cal_cor(add_abund_table = pro.func$res_spe_func_perc, cor_method = "spearman",
           p_adjust_method = "BH", p_adjust_type = "All")
t3$plot_cor(pheatmap = TRUE)
t3$plot_cor(pheatmap = TRUE,filter_feature = c(""))

```


```{r, fig.height=5.5,fig.width=8}

t3_tab <- t3$res_cor


t3_tab_sig <- t3_tab[t3_tab$AdjPvalue < 0.05, ]
t3_tab_sig

t3_tab_sig2 <- t3_tab %>%
  mutate(Correlation2 = case_when(
      AdjPvalue >= 0.05 ~ NA,
      TRUE ~ Correlation)) %>%
  dplyr::filter(Taxa != "ureolysis") %>%
  dplyr::filter(Taxa != "plastic_degradation") %>%
  dplyr::filter(!str_detect(Taxa, "human_")) %>%
  dplyr::filter(Taxa != "cellulolysis") %>%
  dplyr::filter(!str_detect(Taxa, "arsenate")) %>%
  dplyr::filter(!str_detect(Taxa, "parasite")) %>%
  dplyr::filter(!str_detect(Taxa, "aromatic")) %>%
  dplyr::filter(!str_detect(Taxa, "hydrocarbon")) %>%
  dplyr::filter(Env != "Sample_type3") %>%
  dplyr::filter(Env != "V")

#t3$res_cor[ which(t3$res_cor$Taxa == c("methanotrophy", "methanol_oxidation", "nitrification")),]

func_heat1 <- ggplot(t3_tab_sig2, aes(Env, Taxa, fill= Correlation2)) + 
  geom_tile() + theme_bw() +
  scale_fill_gradient(low = "navy", high = "red", na.value="white",
                      name = "Correlation") +
  theme(axis.text.x = element_text(angle=90, vjust=0.5,hjust=1),
        legend.position = "bottom",
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        panel.grid = element_blank())

func_heat1

t3_tab_sig3 <- t3_tab %>%
  dplyr::mutate(Correlation2 = case_when(
      AdjPvalue >= 0.05 ~ NA,
      TRUE ~ Correlation)) %>%
  dplyr::filter(Taxa != "ureolysis") %>%
  dplyr::filter(Taxa != "plastic_degradation") %>%
  dplyr::filter(!str_detect(Taxa, "human_")) %>%
  dplyr::filter(Taxa != "cellulolysis") %>%
  dplyr::filter(!str_detect(Taxa, "arsenate")) %>%
  dplyr::filter(!str_detect(Taxa, "parasite")) %>%
  dplyr::filter(!str_detect(Taxa, "aromatic")) %>%
  dplyr::filter(!str_detect(Taxa, "hydrocarbon")) %>%
  dplyr::filter(Env == "Sample_type3")

t3_tab_sig3

t3_tab_sig3$Env <- as.character(t3_tab_sig3$Env)

t3_tab_sig3[t3_tab_sig3 == "Sample_type3"] <- "Sample Category"

func_heat2 <- ggplot(t3_tab_sig3, aes(Env, Taxa, fill= Correlation2)) + 
  geom_tile() + theme_bw() +
  theme(axis.text.x = element_text(angle=90, vjust=0.5,hjust=1),
        legend.position = "bottom",
        axis.title.x = element_blank(),
        panel.grid = element_blank()) +
  ylab("Functional Category") + 
  scale_fill_gradient(high = "#F9A620", low = "#67a9cf", na.value="white",
                      breaks=c(-0.8, 0.8), name = "",
                      limits=c(-0.8, 0.8), labels=c("Subg.","Exp."))

func_heat2

pp <- list(func_heat2, func_heat1)
func_heatf <- cowplot::plot_grid(plotlist=pp, ncol=2, align='h')
func_heatf

ggsave("functional_heatmap2.svg",
       func_heatf, width = 180, height = 180, units = "mm", dpi = 300)

```


# Explore data further

What are the ASVs responsible for the significant profiles?


```{r}

tab_test <- pro.func$res_spe_func



rowSums(tab_test["ASV_1",]) # The most abundant Crenarchiota ASV has no assigned functions
rowSums(tab_test["ASV_2",]) # 0
rowSums(tab_test["ASV_3",]) # 0
rowSums(tab_test["ASV_4",]) # 0
rowSums(tab_test["ASV_5",]) # 1 ! Proteo
rowSums(tab_test["ASV_6",]) # 2 ! Cyanobacteria
rowSums(tab_test["ASV_7",]) # 0
rowSums(tab_test["ASV_8",]) # The most abundant WPS ASV has no assigned functions
rowSums(tab_test["ASV_9",]) # The most abundant Chloroflexi
rowSums(tab_test["ASV_10",]) # 2 Cyanobacteria
rowSums(tab_test["ASV_11",]) # 0
rowSums(tab_test["ASV_12",]) # 1 Cyanobacteria
rowSums(tab_test["ASV_13",]) # 0: Acido
rowSums(tab_test["ASV_14",]) # 0: ACido
rowSums(tab_test["ASV_15",]) # The most abundant Actinobacteria has no function assigned
rowSums(tab_test["ASV_16",]) # The most abundant GAL15 has no function assigned

```


```{r}
# Use this code to find the ASVs responsible for a certain functional profile that correlate significantly with any of the investigated environmental factors and/or sample type

subset_rows <- tab_test[tab_test$methylotrophy == 1, ]
subset_rows
```


```{r}
phy_all_trans <- transform_sample_counts(phy_all, function(x) x / sum(x))


```


How many ASVs could be assigned a function?

```{r}
# Count the number of rows with at least one 1 in any column
num_rows_with_1 <- sum(apply(tab_test, 1, function(row) any(row == 1)))
num_rows_with_1

num_rows_with_1
668/2734*100
# 24.43307 % could be assigned a function

tab_test_no_zeros <- tab_test[rowSums(tab_test != 0) > 0, ]
tab_test_no_zeros


```

How much does these ASVs constitute of the samples?

```{r}
# Assume 'physeq' is your original phyloseq object
# Specify the ASV names you want to keep in a vector
selected_asvs <- rownames(tab_test_no_zeros)


# Subset the phyloseq object to include only specific ASVs with known function
physeq_subset <- subset_taxa(phy_all_trans, taxa_names(phy_all_trans) %in% selected_asvs)
physeq_subset

```
```{r, fig.height=10, fig.width=10}

erebus_p_physeq_subset <- physeq_subset %>% psmelt() 

erebus_p_physeq_subset$Phylum[is.na(erebus_p_physeq_subset$Phylum)] = "Unassigned"

# Plot 
plot_physeq_subset<- ggplot(erebus_p_physeq_subset, aes(x = Name, y = Abundance, fill = Phylum)) +
  geom_bar(stat = "identity") +theme_bw()+
  theme(
    strip.background = element_blank(),
    strip.text.x = element_text(size = 16),
    panel.background = element_blank(),
    axis.ticks.x = element_blank(),
    axis.text.x  = element_text(size=16, vjust = 0.5, hjust = 1, angle = 90),
    axis.title.x = element_blank(),
    axis.text.y  = element_text(size=16),
    axis.title.y = element_text(size=16),
    plot.title = element_text(size=20, hjust = -0.05),
    legend.text = element_text(size=15), 
    legend.title = element_text(size=15, angle = 90),
    legend.position = "bottom") +
  coord_cartesian(expand = FALSE)

plot_physeq_subset <- plot_physeq_subset + facet_grid(~Sample_type4, scales = "free", space = "free")

#plot_Class$data$Sample_type3 <- factor(plot_Class$data$Sample_type3, levels = c("Subglacial", "Exposed"))
plot_physeq_subset$data$Sample_type4 <- factor(plot_physeq_subset$data$Sample_type4, levels = c("Subglacial", "Exp Hot Soil", "Western Crater", "Tramway Ridge"))

plot_physeq_subset

```

Calculate how much these ASVs comprise the communities to get exact numbers and averages

```{r}
data_abund <- plot_physeq_subset$data

data_abund

#Calculate how much the ASVs with known function constitute in the samples on average across the whole dataset

sum(data_abund$Abundance)/47*100   # 21.83749

#Calculate averages from each group
library(reshape2)
melted_data <- melt(data_abund,  id.vars= "Sample_type4" , measure.vars= "Abundance", variable_name = "Sample")

melted_data_subglacial <- melted_data[which(melted_data$Sample_type4 == "Subglacial"),]
melted_data_exp <- melted_data[which(melted_data$Sample_type4 == "Exp Hot Soil"),]
melted_data_TR <- melted_data[which(melted_data$Sample_type4 == "Tramway Ridge"),]

sum(melted_data_subglacial$value)/9*100 #24.85584
sum(melted_data_exp$value)/17*100 #19.75987
sum(melted_data_TR$value)/21*100 #22.2258

#Calculate the sum of the sample with least ASVs with known function
melted_data_low <- data_abund[which(data_abund$Sample == "Exp Hot Soil 7"),]
melted_data_low
sum(melted_data_low$Abundance)*100   #4% !!!!

```





```{r}
plot_physeq_subset$data
```


